Lost in Translation: \\ The function--sheaf dictionary for smooth commutative group schemes over finite fields

### 2014-04-09 dictionary.tex
Commit before meeting with David.

%%%%%%%%%%%%%%%%% PACKAGES %%%%%%%%%%%%%%%%%
% Encodings
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
% AMS packages
\usepackage{amsmath, amsthm, amssymb, todonotes, showkeys}
\usepackage[alphabetic]{amsrefs}
% Fonts
\usepackage{mathrsfs, yfonts}
\usepackage{geometry}
% TikZ
\usepackage{tikz}
\usetikzlibrary{shapes,arrows,calc,matrix}
\usepackage{tikz-cd}
% Hyperrefs
\usepackage{hyperref}

@@

\subsection*{Overview}
Deligne's function--sheaf dictionary translates characters of connected, commutative algebraic groups $G$ over finite fields $\Fq$ into certain local systems on G, from which the character of $G(\Fq)$ can be recovered using the action of Frobenius on $G$.
The local systems which appear in this dictionary are easily characterised and often referred to as \emph{character sheaves} on $G$.
The result is a perfect dictionary between characters of $G(\Fq)$ and character sheaves on G, with which the formidable artillery of algebraic geometry can be brought to bear on the study of characters of $G(\Fq)$.
Although this dictionary extends to the case when $G$ is not connected, and even to all smooth commutative group schemes over $\Fq$, the result is not perfect.
In this generality, the correct class of sheaves on G are called \emph{quasicharacter sheaves}; when $G$ is connected and of finite type, quasicharacter sheaves on $G$ specialize to character sheaves on $G$.
While every character of $G(\Fq)$ may be converted, canonically, into a quasicharacter sheaf on G from which the character of $G(\Fq)$ may be recovered using the trace of Frobenius on G, generally there exist non-trivial quasicharacter sheaves on G for which the trace of Frobenius is trivial; these are called \emph{invisible quasicharacter sheaves} on $G$.
Invisible quasicharacter sheaves on G are precisely the quasicharacter sheaves on $G$ that are lost in translation under the function-sheaf dictionary.
Assuming only that $G$ is a smooth commutative group scheme over $\Fq$ with finitely generated component group, this paper gives a simple cohomological description of invisible quasicharacter sheaves on $G$. The most interesting case is that of etale commutative group schemes over $\Fq$.

@@

In order to give the main result of this paper with more precision, let us briefly review how the function--sheaf dictionary for connected commutative algebraic groups over finite fields \cite{deligne:SGA4.5}*{sommes trig.} translates characters into local systems.
Let $\Fq$ be a finite field and let $G$ be a connected commutative algebraic group over $\Fq$; fix an algebraic closure $\bFq$ of $\Fq$
and let $\bG$ be the base change of $G$ to $\bFq$.
Fix a prime $\ell$ invertible in $\Fq$ and let $\chi : G(\Fq) \to \EEx$ be any character.
Using little more than the Lang map for $G$, $\chi$ determines an $\ell$-adic local system $\qcs{L}$ on $G$ from which the character $\chi$ can be recovered using the Frobenius automorphism of $\bG$.
While this recovery process applies to all $\ell$-adic local systems $\qcs{L}$ on $G$, those that produce characters of $G(\Fq)$ are distinguished by a simple property: there exists an isomorphism
\begin{equation}\label{introbox}
m^* \qcs{L} \iso \qcs{L} \boxtimes \qcs{L}
\end{equation}
 of $\ell$-adic local systems on $G$, where $m : G\times G \to G$ is the multiplication map for $G$. The function--sheaf dictionary for $G$ identifies a monoidal category of $\ell$-adic local systems on $G$ such that the group of isomorphism classes in this category is canonically identified with the $\ell$-adic character group of $G(\Fq)$, thus providing the geometrization and categorification of characters of $G(\Fq)$.

Although hidden in the discussion above, the connectness of $G$ played a crucial role in the geometrization and categorification of characters of $G(\Fq)$.  So it is natural to ask how the function--sheaf dictionary generalizes to not-necessarily connected commutative algebraic groups over finite fields. In this paper we answer that question by exhibiting a function--sheaf dictionary, in the sense above, for any smooth commutative group scheme $G$ over a finite field $\Fq$.
To do this we must pass from $\ell$-adic local systems on $G$ to Weil local systems on $G$
and also keep track of the isomorphisms \eqref{introbox} (mere existence of such an isomorphism is not enough), leading to the notion of {\it quasicharacter sheaves on $G$}.
But the result  . . . .

So, what is a quasicharacter sheaf?
A quasicharacter sheaf on $G$ is a triple $\qcs{L}\ceq
(\gqcs{L},\mu,\phi)$ where $\gqcs{L}$ is an $\ell$-adic local system on $\bG$ of rank~one and $\mu: m^*
\gqcs{L} \to \LxL$ and $\phi : \Frob{G}^*\gqcs{L} \to \gqcs{L}$ are isomorphisms satisfying natural compatibilty conditions; here $\Frob{G}$ is the Frobenius automorphism of ${\bar G}$.
The main result of this paper is the following statement (see Theorem~\ref{thm:snake}):
the category $\QC(G)$ of quasicharacter sheaves on $G$
is a rigid monoidal category and Trace of Frobenius provides a functorial surjection
\begin{equation}
\QCiso{G} \to \Hom(G(\Fq), \EEx)
\end{equation}
with a canonical section,
and the kernel of the Trace of Frobenius is classified by $\Hh^1(\Weil{},\Hh^2(G(\bFq),\EEx))$.
In particular, in contrast to the case of connected commutative algebraic groups, the Trace of Frobenius may not injective.
We refer to quasicharacter sheaves in the kernel of Trace of Frobenius as \emph{invisible} quasicharacter sheaves.
The canonical section of the Trace of Frobenius allows one to convert any quasicharacter of $G(\Fq)$
into a quasicharacter sheaf on $G$.
In this way, quasicharacter sheaves geometrize characters of $G(\Fq)$.

@@

We close the introduction with a more detailed description of the structure of this paper.
After setting some notation in Section~\ref{sec:QCschemes}, in Section~\ref{sec:category} we define quasicharacter sheaves and give an interpretation of
them as Weil sheaves equipped with an isomorphism of the form  \eqref{introbox}.
Section~\ref{sec:Frob} introduces Trace of Frobenius,
one direction of the function--sheaf dictionary.
We define pullbacks and products of quasicharacter sheaves in Section~\ref{sec:pullback}, and use
them to prove a functoriality result on Trace of Frobenius.
In Section~\ref{sec:basechange} we study how quasicharacter sheaves behave under base change
and Weil restriction.
Sections~\ref{sec:bounded} and \ref{sec:finite} describe bounded quasicharacter sheaves and finite quasicharacter sheaves, respectively,
forming two important subcategories of $\QC(G)$.
%These subcategories play a key role in the proof of Proposition~\ref{prop:restriction}.
In Section \ref{sec:connected} we show that all quasicharacter sheaves on $G$ are finite quasicharacter sheaves when $G$ is connected, and that Trace of Frobenius is an isomorphism in this case.
We shift to \'etale group schemes $G$ in Section~\ref{sec:etale} and give a more concrete interpretation
of quasicharacter sheaves in this case.  Using a notion of global sections for quasicharacter sheaves, in this section
we also give isomorphisms between $\QCiso{G}$, $\Hh^1(\Weil{}, \Hom(G(\bFq), \EEx))$ and $\Hom(G(\Fq), \EEx)$, where $\Weil{}$ is the Weil group for $\Fq$.
In Section~\ref{sec:restriction} we show that for non-connected $G$, restriction to the identity component
induces an essentially surjective functor of quasicharacter sheaves.
Section~\ref{sec:snake} presents the main theorem of this paper: Trace of Frobenius
is a functorial isomorphism for $\QC$-schemes.

@@

Throughout this paper, $G$ is a smooth, commutative group scheme
over a finite field $\Fq$, and $m : \GxG\to G$ is its multiplication morphism.  For the sake of brevity, we
define a \emph{$\QC$-scheme} to be such a $G$ with the additional property that the
geometric component group of $G$ is finitely generated.  This last condition is only
needed in the proof of Lemma \ref{lemma:section}.  A morphism of $\QC$-schemes is a morphism
of group schemes over $\Fq$.

@@

We fix a prime $\ell$, invertible in $\Fq$.
We will work with constructible $\ell$-adic sheaves \citelist{\cite{deligne:80a}*{\S 1.1} \cite{SGA5}*{Expos\'es V, VI}}
on $\QC$-schemes, employing the standard formalism.
We also make extensive use of the external tensor product of $\ell$-adic sheaves,
defined as follows: if $\mathcal{F}$ and $\mathcal{G}$ are constructible $\ell$-adic
sheaves on schemes $X$ and $Y$ and $p_X : X\times Y\to X$ and $p_Y : X\times Y \to Y$
are the projections, then $\mathcal{F}\boxtimes \mathcal{G} \ceq p_X^* \mathcal{F} \otimes p_Y^*\mathcal{G}$.

@@

We will see that the Trace of Frobenius may not provide complete
information about isomorphism classes of quasicharacter sheaves on $G$ when $G$ is not connected.
Our main tool for understanding this phenomenon is the function
\begin{equation}\label{2cocycle}
S_{G}: \QCiso{G} \to H^0(\Weil{},H^2(G(\bFq),\EEx))
\end{equation}
defined here.
Let $\qcs{L} = (\gqcs{L},\mu,\phi)$ be a quasicharacter sheaf on $G$.
For each geometric point $x\in G(\bFq)$, choose a basis $\{ v_x \}$ for $\gqcs{L}_x$.
This choice determines a function $g : G(\bFq)\times G(\bFq) \to \EEx$ defined by
\[
\mu_{x,y}(v_{x+y}) = g(x,y) v_x \otimes v_y.
\]
Chasing through the diagram in Condition~\ref{QC.2} we see that
\[
g(x+y,z) g(x,y) = g(x,y+z) g(y,z),
\]
for all $x,y,z\in G(\bFq)$, so $g \in Z^2(G(\bFq),\EEx)$.
It is easy to see that the class of $g$ is independent of the choice made above;
let $g_\mathcal{L} \in H^2(G(\bFq),\EEx)$ the class of $g\in Z^2(G(\bFq),\EEx)$.
It is also easy to see that $g_\mathcal{L} = g_\mathcal{L'}$ when $\qcs{L} \iso \qcs{L'}$. This defines
\[
\QCiso{G} \to H^2(G(\bFq),\EEx).
\]
We claim that $g_\qcs{L}$ is also fixed by the action of $\Weil{}$ on $H^2(G(\bFq),\EEx)$
inherited from the action of $\Weil{}$ on $G(\bFq)$.
To see this, we return to the choice $\{ v_x \in \gqcs{L}_x^\times \tq x\in G(\bFq) \}$
and define $f : G(\bFq) \to \EEx$ by
\[
\phi_x(v_{\Frob{G}(x)}) = f(x) v_x.
\]
Chasing through the diagram in Condition~\ref{QC.3} we see that
\begin{equation}\label{nohom}
g(\Frob{G}(x),\Frob{G}(y)) f(x) f(y)  = g(x,y) f(x+y)
\end{equation}
for all $x,y \in G(\bFq)$; thus, the class of $\,^{\Frob{}}g$ coincides with the class of $g$.
This concludes the definition of
$S_{G}: \QCiso{G} \to H^0(\Weil{},H^2(G(\bFq),\EEx))$.
It is easy to see that the function $S_{G}: \QCiso{G} \to H^0(\Weil{},H^2(G(\bFq),\EEx))$ is a group homomorphism:
\[
g_{\qcs{L}\otimes \qcs{L'}} = g_{\qcs{L}} \, g_\qcs{L'}
\]
for every $\qcs{L}, \qcs{L'} \in \QC(G)$.\todo{I still need to write down the details, for ourselves.}

>>>>

%%%%%%%%%%%%%%%%% PACKAGES %%%%%%%%%%%%%%%%%
% Encodings
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
% AMS packages
\usepackage{amsmath, amsthm, amssymb, todonotes}% showkeys}
\usepackage[alphabetic]{amsrefs}
% Fonts
\usepackage{mathrsfs, yfonts}
\usepackage{geometry}
% TikZ
\usepackage{tikz}
\usetikzlibrary{shapes,arrows,calc,matrix}
\usepackage{tikz-cd}
% Hyperrefs
\usepackage{hyperref}

@@

\subsection*{Overview}
Deligne's function--sheaf dictionary translates characters of connected, commutative algebraic groups $G$ over finite fields $\Fq$ into certain local systems on G, from which the character of $G(\Fq)$ can be recovered using the action of Frobenius on $G$.
The local systems which appear in this dictionary are easily characterised and often referred to as \emph{character sheaves} on $G$.
The result is a perfect dictionary between characters of $G(\Fq)$ and character sheaves on G, with which the formidable artillery of algebraic geometry can be brought to bear on the study of characters of $G(\Fq)$.
Although this dictionary extends to the case when $G$ is not connected, and even to all smooth commutative group schemes over $\Fq$, the result is not perfect.
In this generality, the correct class of sheaves on G are called \emph{quasicharacter sheaves}; when $G$ is connected and of finite type, quasicharacter sheaves on $G$ specialize to character sheaves on $G$.
While every character of $G(\Fq)$ may be converted, canonically, into a quasicharacter sheaf on G from which the character of $G(\Fq)$ may be recovered using the trace of Frobenius on G, generally there exist non-trivial quasicharacter sheaves on G for which the trace of Frobenius is trivial; these are called \emph{invisible quasicharacter sheaves} on $G$.
Invisible quasicharacter sheaves on G are precisely the quasicharacter sheaves on $G$ that are lost in translation under the function-sheaf dictionary.
Assuming only that $G$ is a smooth commutative group scheme over $\Fq$ with finitely generated component group, this paper gives a simple cohomological description of invisible quasicharacter sheaves on $G$. 
%The most interesting case is that of etale commutative group schemes over $\Fq$.

@@

In order to describe the main result of this paper with more precision, let us briefly review how the function--sheaf dictionary for connected commutative algebraic groups over finite fields \cite{deligne:SGA4.5}*{sommes trig.} translates characters into local systems.
Let $\Fq$ be a finite field and let $G$ be a connected commutative algebraic group over $\Fq$; fix an algebraic closure $\bFq$ of $\Fq$
and let $\bG$ be the base change of $G$ to $\bFq$.
Fix a prime $\ell$ invertible in $\Fq$ and let $\chi : G(\Fq) \to \EEx$ be any character.
Using little more than the Lang map for $G$, $\chi$ determines an $\ell$-adic local system $\qcs{L}$ on $G$ from which the character $\chi$ can be recovered using the Frobenius automorphism of $\bG$.
While this recovery process applies to all $\ell$-adic local systems $\qcs{L}$ on $G$, those that produce characters of $G(\Fq)$ are distinguished by a simple property: there exists an isomorphism
\begin{equation}\label{introbox}
m^* \qcs{L} \iso \qcs{L} \boxtimes \qcs{L}
\end{equation}
 of $\ell$-adic local systems on $G$, where $m : G\times G \to G$ is the multiplication map for $G$. The function--sheaf dictionary for $G$ identifies a monoidal category of $\ell$-adic local systems on $G$ such that the group of isomorphism classes in this category is canonically identified with the $\ell$-adic character group of $G(\Fq)$, thus providing the geometrization and categorification of characters of $G(\Fq)$.

Although hidden in the discussion above, the connectness of $G$ played a crucial role in the geometrization and categorification of characters of $G(\Fq)$.  So it is natural to ask how the function--sheaf dictionary generalizes to not-necessarily connected commutative algebraic groups over finite fields. In this paper we answer that question by exhibiting a function--sheaf dictionary, in the sense above, for any smooth commutative group scheme $G$ over a finite field $\Fq$.
To do this we must pass from $\ell$-adic local systems on $G$ to Weil local systems on $G$
and also keep track of the isomorphisms \eqref{introbox} (mere existence of such an isomorphism is not enough), leading to the notion of {\it quasicharacter sheaves on $G$}.
%So, what is a quasicharacter sheaf?

A quasicharacter sheaf on $G$ is a triple $\qcs{L}\ceq
(\gqcs{L},\mu,\phi)$ where $\gqcs{L}$ is an $\ell$-adic local system on $\bG$ of rank~one and $\mu: m^*
\gqcs{L} \to \LxL$ and $\phi : \Frob{G}^*\gqcs{L} \to \gqcs{L}$ are isomorphisms satisfying natural compatibilty conditions; here $\Frob{G}$ is the Frobenius automorphism of ${\bar G}$.

The main result of this paper is the following statement (see Theorem~\ref{thm:snake}):
the category $\QC(G)$ of quasicharacter sheaves on $G$
is a rigid monoidal category and Trace of Frobenius provides a functorial surjection
\begin{equation}
\QCiso{G} \to \Hom(G(\Fq), \EEx)
\end{equation}
with a canonical section,
and the kernel of the Trace of Frobenius is classified by $\Hh^1(\Weil{},\Hh^2(G(\bFq),\EEx))$.
In particular, in contrast to the case of connected commutative algebraic groups, the Trace of Frobenius may not injective.
We refer to quasicharacter sheaves in the kernel of Trace of Frobenius as \emph{invisible} quasicharacter sheaves.
The canonical section of the Trace of Frobenius allows one to convert any quasicharacter of $G(\Fq)$
into a quasicharacter sheaf on $G$.
In this way, quasicharacter sheaves geometrize characters of $G(\Fq)$.

@@

%We close the introduction with a more detailed description of the structure of this paper.
After setting some notation in Section~\ref{sec:QCschemes}, in Section~\ref{sec:category} we define quasicharacter sheaves and give an interpretation of
them as Weil sheaves equipped with an isomorphism of the form  \eqref{introbox}.
Section~\ref{sec:Frob} introduces Trace of Frobenius,
one direction of the function--sheaf dictionary.
We define pullbacks and products of quasicharacter sheaves in Section~\ref{sec:pullback}, and use
them to prove a functoriality result on Trace of Frobenius.
In Section~\ref{sec:basechange} we study how quasicharacter sheaves behave under base change
and Weil restriction.
Sections~\ref{sec:bounded} and \ref{sec:finite} describe bounded quasicharacter sheaves and finite quasicharacter sheaves, respectively,
forming two important subcategories of $\QC(G)$.
%These subcategories play a key role in the proof of Proposition~\ref{prop:restriction}.
In Section \ref{sec:connected} we show that all quasicharacter sheaves on $G$ are finite quasicharacter sheaves when $G$ is connected, and that Trace of Frobenius is an isomorphism in this case.
We shift to \'etale group schemes $G$ in Section~\ref{sec:etale} and give a more concrete interpretation
of quasicharacter sheaves in this case.  Using a notion of global sections for quasicharacter sheaves, in this section
we also give isomorphisms between $\QCiso{G}$, $\Hh^1(\Weil{}, \Hom(G(\bFq), \EEx))$ and $\Hom(G(\Fq), \EEx)$, where $\Weil{}$ is the Weil group for $\Fq$.
In Section~\ref{sec:restriction} we show that for non-connected $G$, restriction to the identity component
induces an essentially surjective functor of quasicharacter sheaves.
Section~\ref{sec:snake} presents the main theorem of this paper: Trace of Frobenius
is a functorial isomorphism for $\QC$-schemes.

@@

Throughout this paper, $G$ is a smooth, commutative group scheme
over a finite field $\Fq$, and $m : \GxG\to G$ is its multiplication morphism.  
%For the sake of brevity, we define a \emph{$\QC$-scheme} to be such a $G$ with the additional property that the geometric component group of $G$ is finitely generated.  This last condition is only needed in the proof of Lemma \ref{lemma:section}.  A morphism of $\QC$-schemes is a morphism of group schemes over $\Fq$.

@@

We fix a prime $\ell$, invertible in $\Fq$.
We will work with constructible $\ell$-adic sheaves \citelist{\cite{deligne:80a}*{\S 1.1} \cite{SGA5}*{Expos\'es V, VI}}
on schemes locally of finite type over $\Fq$, employing the standard formalism.
We also make extensive use of the external tensor product of $\ell$-adic sheaves,
defined as follows: if $\mathcal{F}$ and $\mathcal{G}$ are constructible $\ell$-adic
sheaves on schemes $X$ and $Y$ and $p_X : X\times Y\to X$ and $p_Y : X\times Y \to Y$
are the projections, then $\mathcal{F}\boxtimes \mathcal{G} \ceq p_X^* \mathcal{F} \otimes p_Y^*\mathcal{G}$.

@@

We will see that the Trace of Frobenius may not provide complete
information about isomorphism classes of quasicharacter sheaves on $G$ when $G$ is not connected.
Our main tool for understanding this phenomenon is the function
\begin{equation}\label{2cocycle}
S_{G}: \QCiso{G} \to H^0(\Weil{},H^2(G(\bFq),\EEx))
\end{equation}
defined here.
Let $\qcs{L} = (\gqcs{L},\mu,\phi)$ be a quasicharacter sheaf on $G$.
For each geometric point $x\in G(\bFq)$, choose a basis $\{ v_x \}$ for $\gqcs{L}_x$.
This choice determines a function $g : G(\bFq)\times G(\bFq) \to \EEx$ defined by
\[
\mu_{x,y}(v_{x+y}) = g(x,y) v_x \otimes v_y.
\]
Chasing through the diagram in Condition~\ref{QC.2} we see that
\[
g(x+y,z) g(x,y) = g(x,y+z) g(y,z),
\]
for all $x,y,z\in G(\bFq)$, so $g \in Z^2(G(\bFq),\EEx)$.
It is easy to see that the class of $g$ is independent of the choice made above;
let $\sigma_\mathcal{L}) \in H^2(G(\bFq),\EEx)$ the class of $g\in Z^2(G(\bFq),\EEx)$.
It is also easy to see that $S_G(\mathcal{L}) = S_G(\mathcal{L'})$ when $\qcs{L} \iso \qcs{L'}$. This defines
\[
S_G: \QCiso{G} \to H^2(G(\bFq),\EEx).
\]
We claim that $S_G(\qcs{L})$ is also fixed by the action of $\Weil{}$ on $H^2(G(\bFq),\EEx)$
inherited from the action of $\Weil{}$ on $G(\bFq)$.
To see this, we return to the choice $\{ v_x \in \gqcs{L}_x^\times \tq x\in G(\bFq) \}$
and define $f : G(\bFq) \to \EEx$ by
\[
\phi_x(v_{\Frob{G}(x)}) = f(x) v_x.
\]
Chasing through the diagram in Condition~\ref{QC.3} we see that
\begin{equation}\label{nohom}
g(\Frob{G}(x),\Frob{G}(y)) f(x) f(y)  = g(x,y) f(x+y)
\end{equation}
for all $x,y \in G(\bFq)$; thus, the class of $\,^{\Frob{}}g$ coincides with the class of $g$.
This concludes the definition of
$S_{G}: \QCiso{G} \to H^0(\Weil{},H^2(G(\bFq),\EEx))$.
It is easy to see that the function $S_{G}: \QCiso{G} \to H^0(\Weil{},H^2(G(\bFq),\EEx))$ is a group homomorphism:
\[
g_{\qcs{L}\otimes \qcs{L'}} = g_{\qcs{L}} \, g_\qcs{L'}
\]
for every $\qcs{L}, \qcs{L'} \in \QC(G)$.\todo{I still need to write down the details, for ourselves.}
